---
title: "HW6"
author: "Nathan Krieger"
date: "2026-03-06"
output: pdf_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Problem 1: Concept Review

## (a)

Best subset selection will produce a collection of p models M1, M2, . . . , Mp. These represent
the ‘best’ model of each size (where ‘best’ here is defined as the model with the smallest RSS).

### (i) True or False

Is it true that the model identified as Mk+1 must contain a subset of the predictors found
in Mk? In other words, is it true that if M1 : Y ~ X1, then M2 must also contain X1.
And if M2 contains X1 and X2, then M3 must also contain X1 and X2?

## (b)

Same question as part (a) but instead of best subset selection, we now carry out forward
selection.

## (c) (practice)

Suppose we perform best subset selection, forward selection, and backward selection on a training data set.

Best subset selection produces a model with k predictors for each k = 1, . . . , p. Denote
this model by Mk,subset. For example, when k = 5, M5,subset is the best 5-predictor model
according to best subset selection.

Similarly, forward selection produces a model with k predictors, denoted by Mk,forward, and
backward selection produces a model with k predictors, denoted by Mk,backward.

For each case below, state which of the three models would be expected to have the smallest
training MSE, and briefly explain why.

### (i) (practice) k = 1

Solution: For k = 1, subset has the same model as forward stepwise and therefore the
same RSS (and training MSE). Backward stepwise may have a different model, and a
worse RSS.

### (ii) (practice)k ${\epsilon}$ {2, ..., p - 2}

### (iii) (practice)k = p - 1

### (iv) (practice)k = p


## (d) (practice)

Same setup as part (c). For a given k, which of these three models has the smallest
test MSE? Explain your answer.

### (i) k = 1

Solution: Subset and forward stepwise have the same model when k = 1, so they have
the same test MSE. Backward selection may produce a different model, so its test MSE
may differ.

### (ii) k = p

###(iii) k = p - 1

# Problem 2: Simulation Studies

We have seen that as the number of features used in a model increases, the training error will
necessarily decrease, but the test error may not. We will now explore this in a simulated data set.
Start by generating a data set with p = 20 features, n = 1000 observations, and an associated
response vector Y generated according to the model:

Y = $\beta$0 + $\beta$1X1 + $\beta$2X2 + . . . $\beta$pXp + $\epsilon$,

You can use this code to get started:
```{r}
p = 20
n = 1000
error = rnorm(n,0,1)
beta = c(2, rep(0,5),1.2,3,-2,1.5,7,rep(0,5),3,4,1,9,2)
Xmat = matrix(rnorm(n*20),n,20)
X = cbind(rep(1,n),Xmat)
Y = X%*%beta + error
data = data.frame(Y,Xmat)
```

Note that X%*%beta is just doing matrix multiplication: beta0 + beta1*X1 + beta2*X2 + ....

beta20*X20.

## (a) 

Split your data set into a training set containing 100 observations and a test set containing
900 observations.

## (b) 

Perform best subset selection on the training set only, and plot the training set MSE associated
with the best model of each size. Hint: The training MSE can be computed from the model
output. If the fitted model is stored as fit, you can extract it using mean(fit$rss).

## (c) 

Plot the test MSE associated with the best model of each size.

## (d) 

For which model size does the test MSE take on its minimum values? Comment on your
results.

## (e) 

How does the model at which the test MSE is minimized compare to the true model used
to generate the data? Comment on how well the estimated regression coefficients match the
true values.

Note: The true model contains only 10 predictors with nonzero coefficients. The remaining
10 predictors have true coefficient value equal to zero.


# Problem 3: Forward and backward selection (practice)

We will use the College data set in the ISLR2 library to predict the number of applications (Apps)
each university received. Use all remaining variables as predictors.

## (a) (practice)

Randomly split the data set so that 90% of the data belong to the training set
and the remaining 10% belong to the test set. Implement forward and backward selection on
the training set only. Report the test MSE for all models (M1, M2, . . . , Mp) based on forward
selection. Which model minimizes test MSE?

Report the test MSE for all models (M1, M2, . . . , Mp) based on backward selection. Which
model minimizes test MSE?

Do both algorithms lead you to the same model based on test MSE?

## (b) (practice)

Run forward and backward selection on the full dataset. For each approach, report
the best model based on AIC. Do they lead you to the same model based on AIC?

## (c) (practice)

Which is the final model that you would pick? If both (a) and (b) lead you to the
same model, then this is easy. Otherwise, there is no absolute right answer. Try to justify
your choice using some ideas/concepts you’ve learned from this class.


# Problem 4: A Puzzling Problem

When fitting a linear regression model on a data set, you encounter the following R output. You
notice there is something strange about the results. Point out what is strange in this output and
explain clearly how this could happen.

Call:
lm(formula = y ~ x1 + x2)


# Problem 5: Interaction Terms

We will use the Credit dataset for this problem. It is part of the library ISLR2

## (a) 

This data set contains a few categorical predictors. As we already discussed in lecture, these
predictors should be stored as factors so that R can handle them properly. Using the str
function, check that all the qualitative predictors in our dataset are stored correctly in R as
factors. Copy and paste your output.

## (b) 

Fit a model with the response (Y ) as credit card balance and X1= Income and X2 = Student
as the predictors. Call this model fit. Summarize your output.

## (c)

Based on our results from part (b), write out the fitted model for

### i students

### ii non-students.

## (d)

Interpret the regression coefficient related to Income for both models.

### i students

### ii non-students.

## (e) (practice)

Notice that our model says that regardless of student status, the effect of Income
on average Balance is the same. Do you think this is a reasonable constraint on our model?

Construct some plots to back up your answer.

##(f) (practice)

One way we could relax this assumption is by incorporating interaction terms into
our model. Specifically:

Yi = $\beta$0 + $\beta$1Xi1 + $\beta$2Xi2 + $\beta$3Xi3 + $\epsilon$i,

where X1= Income, X2 = Student, and X3 = Income × Student. Fit a model with an
interaction term using the following code:

```{r}
# lm(Balance ~ Income + Student + Income:Student, data=Credit)
```

Based on this model, write out the fitted model for students and write out the fitted model
for non-students.

## (g) (practice)

Interpret the regression coefficient related to Income for the fitted models obtained
in part (f).
