---
title: "HW3"
author: "Nathan Krieger"
date: "2026-02-06"
output: pdf_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

``` {r}

library(ISLR2)
#head(Carseats)

```

# Problem 1

## (a)
### (i)

```{r}

head(Carseats)


m1 <- lm(Sales ~ Sales + CompPrice + Income +Advertising + Population + Price + Age + Education + Urban + US, data=Carseats)

summary(m1)$coefficients

#summary_table <- summary(m1)$coefficients
#print(summary_table)


```



### (ii)

Null hypothesis (H0): The CompPrice of the car seat has zero effect on the sale of the car seats

Alternative hypothesis (H1): The CompPrice of the car seat has an effect on the sale of the car seats

Test statistic: 
``` {r}

comp_price_t_val <- summary(m1)$coefficients["CompPrice", "t value"]

print(comp_price_t_val)

```


```{r}

comp_price_p_val <- summary(m1)$coefficients["CompPrice", "Pr(>|t|)"]

print(comp_price_p_val)

```



Since the p-value is less than $\alpha=0.05$ (2.131214e-28), we reject the null hypothesis and conclude that CompPrice has a statistically significant relationship with car seat Sales.

# TODO: NULL DISTRIBUTION

## (b)

I needed to assume the $\epsilon$ (errors) are normally distributed.

## (c)

```{r}

summary(m1)$sigma^2

```

$\hat{\sigma}^2 = 3.732624$ 

This is the estimated variance of the error term. It represents the average squared deviation of the actual sales values from the predicted regression line.

## (d)

```{r}

summary(m1)$coefficients["Advertising", ]

```

## (e)
## (f)
## (g)
## (h)
## (i)
## (j)
## (k)


# Problem 2

## (a)

$m * \alpha$


## (b)

``` {r}

set.seed(123)
n = 1000  # Number of observations
p_test = c(200, 400, 500, 600, 800)
alpha = 0.05

false_positives <- c()

for (p in p_test) {
  x <- matrix(rnorm(n * p), n, p)
  y <- rnorm(n)
  
  data <- as.data.frame(x)
  
  model <- lm(y ~ ., data = data)
  
  #Step 6: Extract p-values for all predictors
  p_values <- summary(model)$coefficients[-1,4]  # use -1 to Remove intercept
  
  #Step 7: Count number of significant predictors at 0.05 level
  significant_count = sum(p_values < alpha)
  # or 
  false_positives <- c(false_positives, significant_count)
}

plot(p_test, false_positives, type="b",
     xlab = "# of tests carried out",
     ylab = "# of of false positives"
)

```

# Problem 3

## (a)

$\hat{\beta}_0 = 2$
$\hat{\beta}_1 = 3$
$\hat{\beta}_2 = 5$

## (b)

```{r}

X1 = seq(0,10,length.out =100) #generates 100 equally spaced values from 0 to 10.
X2 = runif(100) #generates 100 uniform values.

```

## (c)
## (d)
## (e)

# Problem 4

## (a)

In plain language, setting the significance level to a = 0.05 means that before we look at the data, we are deciding that we’re willing to accept a 5% chance of making a false alarm.

## (b)

I would disagree with that claim. This is because not significant is not the same as no effect. 

## (c)

