---
title: "HW7"
author: "Nathan Krieger"
date: "2026-03-23"
output: pdf_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Problem 1: Multiple Linear Regression

Suppose you work for a consulting firm. Your manager gives you the Boston dataset (part of
library ISLR2) and asks you to build a multiple linear regression model to predict median home
prices (medv). In typical form, he gives you no instructions or input. The final deliverable should
be a model that can accurately predict median home price from a set of predictors.

Create a report (no more than 1 page) that includes the following items/discussion:

```{r}

library(ISLR2)
library(leaps)


set.seed(123)


best_model <- regsubsets(medv ~ ., data = Boston, nbest = 1, nvmax = 12)

summary(best_model)
smallest <- which.min(summary(best_model)$bic)

best <- coef(best_model, id = smallest)

```

## (a)

Report your final model and its regression coefficients in a nicely summarized table.


```{r}

knitr::kable(best)

```

TODO: write something

## (b)

Not all predictors need to be in the model. What technique will you use to determine which
predictors to include in the model and which to omit? Explain your workflow and justify how
you decided on your “final” model. Directly address why you choose your final model over
other candidate models.

To find which predictors should be left out of the model, I used best subset selection. 

```{r}

res_sum <- summary(best_model)

par(mfrow = c(1, 2))
plot(res_sum$bic, type = "b", pch = 19, col = "navy", main = "BIC vs Model Size")
points(opt_size, res_sum$bic[opt_size], col = "red", cex = 2, pch = 4, lwd = 2)

plot(best_subset, scale = "bic", main = "Variable Inclusion Grid")

plot(res_sum$cp, type = "b", pch = 19, col = "firebrick", 
     xlab = "Number of Variables", ylab = "Cp (AIC Proxy)",
     main = "AIC/Cp vs Model Size")
points(best_aic, res_sum$cp[best_aic], col = "black", cex = 2, pch = 4, lwd = 2)

plot(res_sum$adjr2, type = "b", pch = 19, col = "darkgreen", 
     xlab = "Number of Variables", ylab = "Adjusted R-squared",
     main = "Adj R^2 vs Model Size")
points(best_r2, res_sum$adjr2[best_r2], col = "red", cex = 2, pch = 4, lwd = 2)

```

## (c) 

What assumptions were needed to obtain this final model? Are there any diagnostic checks
you made to ensure those assumptions were valid? Did you update your final model to
accommodate any assumption violations?

## (d) 

Were there any potential issues related to your final model that you explored? Did you
address these issues once identified?

## (e) 

Can you obtain the bias and variance of this model? If so, explain how. If not, explain why
not.

Since you are the only data scientist on the team, your manager reminds you to make every
effort to produce a report that is rigorous but understandable to non-statisticians. Limita-
tions/strengths of your pipeline to build this model should be discussed. Your workflow should be
thoughtfully justified using the concepts/ideas we have discussed in class.


# Problem 2: Follow-up to in-class activity

Let us recap what we learned from our in-class activity. See the post on Ed Discussion if you need
a refresher.

## (a) 
Clearly the models in Part 2 and Part 3 are not equivalent. Explain again in plain language
(using plots if you deem it helpful) why these two approaches will not give the same results.

Running two separate models assumes that all coefficients are different. Running one model with just a gender dummy variable assumes only the intercept is different while the slopes for other variables stay the same.

## (b) 
Load in the Insurance dataset again. Show how you can modify the model in Part 2 so that
it is exactly equivalent to the models in Part 3. Present your code and results to show that
the fitted models for males only and females only are exactly the same in Part 2 and Part 3
(after making your modification). Hint: use interaction terms.

```{r}

insurance = read.csv('~/git/DS301/HW7/insurance.csv', header = TRUE)

fit <- lm(charges ~ age * gender + bmi * gender, data = insurance)

summary(fit)
coef(fit)

```

## (c) 
Other than domain knowledge, how might you determine using a data-driven way that an
interaction term is needed in the model? Justify and then implement your approach here on
the Insurance dataset.

We can do a partial F-Test.
We compare a Reduced Model to an Augmented Model. If the improvement in the Sum of Squared Errors is statistically significant, the interaction is justified.

```{r}

red <- lm(charges ~ age + bmi + gender, data = insurance)
full <- lm(charges ~ age*gender + bmi*gender, data = insurance)


anova(red, full)

```

Based on my analysis. The p-value of the full model is 0.2607. This is much higher than a = 0.05 and therefore it is not worth the added complexity. 

# Problem 3: Predictions in the presence of multicollinearity (practice)

## (a) 
Is multicollinearity a problem for making accurate predictions? Make an educated guess
based on what we have learned in class.

I would say no. It makes it hard to determine the effect of individual variables, the model remains accurate.

## (b) 
Let us carry out a simulation study to answer this. We will simulate data with and without
multicollinearity. This is our true model:
Yi = B0 + B1Xi1 + B2Xi2 + epsilon_i,

where i = 1, . . . , 100, epsilon ~ N (u = 0, sigma2 = 4), B0 = 3, B1 = 2, and B2 = 4.

To generate your predictors with multicollinearity, use the following code:

```{r}
set.seed(42)
x1 = runif(100)
x2 = 0.8*x1 + rnorm(100,0,0.1)

cor(x1, x2)
```

Using these predictors, generate your Y values in R. Check the correlation between x1 and
x2 using the cor() function. Report that value here.

Correlation is 0.9404249

## (c) 
Split your data into a training set and test set. Train your model on the training set.

```{r}

set.seed(42)

Y = 3 + 2*x1 + 4*x2 + rnorm(100, 0, 2)

df <- data.frame(Y, x1, x2)
head(df)
n <- dim(df)[1]

train_index <- sample(1:n,n/2,rep=FALSE)
train <- df[train_index,]
test <- df[-train_index,]

m <- lm(Y ~ x1 + x2, data = train)

pred <- predict(m, newdata = test)
test_mse <- mean((pred - test$Y)^2)
test_mse

```

Report the test MSE for this model.

Test MSE: 4.683967

## (d) 
Repeat this process 2,500 times. Each time you will need to generate a random set of Y ’s, fit
a model on your training set, and then obtain the test MSE for that model. We do not need
to generate new predictor values. Remember to store the test MSE for each iteration
and do not set a seed within the loop. What is the mean test MSE in this setting when
the predictors are highly correlated? Plot a histogram of your 2,500 test MSEs and comment
on what you observe.

```{r}

set.seed(42)

test_mse <- rep(NA, 2500)
test_mat <- model.matrix(Y ~ ., data = test)

for(i in 1:2500) {
  # New random noise for Y
  y_sim <- 3 + 2*x1 + 4*x2 + rnorm(n, 0, 2)
  
  df <- data.frame(Y, x1, x2)
  
  train_index <- sample(1:n,n/2,rep=FALSE)
  train <- df[train_index,]
  test <- df[-train_index,]
  
  # Fit and Predict
  m <- lm(Y ~ x1 + x2, data = train)
  pred <- predict(m, newdata = test)
  test_mse[i] <- mean((pred - test$Y)^2)
}

hist(test_mse)


```

Based on the plot we can see that the mean is around 4 which is around the sigma squared. 

## (e) 
Now generate predictors without multicollinearity using the following code:

```{r}
set.seed(24)
x1 = runif(100)
x2 = rnorm(100,0,1)

cor(x1, x2)

Y = 3 + 2*x1 + 4*x2 + rnorm(100, 0, 2)

```

Using these predictors, generate your Y values in R. Check the correlation between x1 and
x2. Report that value here.

Correlation: 0.03316596

## (f) 
Again run 2,500 simulations to obtain the test MSE of our model when the predictors are
not correlated. What is the mean test MSE in this setting without multicollinearity?
Plot a histogram of your 2,500 test MSEs and comment on what you see.

```{r}

set.seed(24)

test_mse <- rep(NA, 2500)
test_mat <- model.matrix(Y ~ ., data = test)

for(i in 1:2500) {
  # New random noise for Y
  y_sim <- 3 + 2*x1 + 4*x2 + rnorm(n, 0, 2)
  
  df <- data.frame(Y, x1, x2)
  
  train_index <- sample(1:n,n/2,rep=FALSE)
  train <- df[train_index,]
  test <- df[-train_index,]
  
  # Fit and Predict
  m <- lm(Y ~ x1 + x2, data = train)
  pred <- predict(m, newdata = test)
  test_mse[i] <- mean((pred - test$Y)^2)
}

hist(test_mse)

```

Based on the plot: the mean is the same (~4).

## (g) 
Based on our simulation study, is multicollinearity a problem for making accurate predictions?
Comment on your findings.

No. We can see both test MSEs are about the same. Multicollinearity increases variance but the variance cancels out.



