---
title: "DS 3010 R Coding Cheatsheet Template"
author: ""
date: "Spring 2026"
output:
  html_document:
    toc: true
    toc_depth: 2
    theme: readable
  pdf_document:
    toc: true
    toc_depth: 2
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, eval = FALSE, message = FALSE, warning = FALSE)
```

# How to Use This Sheet

Use this sheet when you are working on a coding problem and are not sure which R function to use. The sections are organized by task. In this template version, most code has been removed so you can practice filling in the correct R commands.

---

# 1. Load Common Packages

| Goal | Package | Code |
|---|---|---|
| ISLR datasets | `ISLR2` | `library(ISLR2)` |
| Linear / quadratic discriminant analysis | `MASS` | `library(MASS)` |
| Naive Bayes | `e1071` | `library(e1071)` |
| Naive Bayes with kernel density | `klaR` | `library(klaR)` |
| KNN | `class` | `library(class)` |
| Cross-validation folds | `caret` | `library(caret)` |
| Classification/regression trees | `tree` | `library(tree)` |
| Bagging and random forests | `randomForest` | `library(randomForest)` |
| Boosting | `gbm` | `library(gbm)` |
| ROC / AUC | `ROCR` | `library(ROCR)` |

```{r load-packages}
library(ISLR2)
library(MASS)
library(e1071)
library(klaR)
library(class)
library(caret)
library(tree)
library(randomForest)
library(gbm)
library(ROCR)
```

---

# 2. Basic Data Setup


## View the data

```{r view-data}
# View the first few rows of the data.

head(Hitters)

# Check the dimensions of the data.

dim(Hitters)
nrow(Hitters)
ncol(Hitters)

# Get summary statistics.

summary(Hitters)

# View the variable names.

str(Hitters)

```

## Remove missing values

```{r remove-missing}
# Remove observations with missing values.

na.omit(Hitters)

```

## Create a new binary response variable

```{r binary-response}
# Create a binary response variable from a quantitative variable.

High <- factor(ifelse(Carseats$Sales <= 8, "No", "Yes"))


```

## Log-transform a response variable

```{r log-transform}

# Create a log-transformed version of a quantitative variable.
X1 = seq(1,100)
X2 = seq(51, 150)
Y = 6+ 4*X1 + 2*X2

df = data.frame(Y, X1, X2)

fit <- lm(log(Y) ~ X1 + X2, data = df)

```

## Remove a variable from the data

```{r remove-variable}
# Remove a variable from the dataset if it should not be used as a predictor.

Carseats$Sales <- NULL

```

---

# 3. Train/Test Split

## Randomly split data into training and test sets

```{r train-test-random}
# Set a seed.

set.seed(1)

# Find the number of observations.

n <- nrow(Hitters)

# Randomly select training row indices.
# Define the test row indices.
index <- sample(1:n,n/2) 

# Create training and test data frames.

train <- Hitters[index, ]
test <- Hitters[-index, ]
```

## Split based on a condition

```{r train-test-condition}
## Split based on a condition

# Create a logical vector for the training set based on a condition.
# Example: Use observations from years before 2005 for training
train_logic <- (Smarket$Year < 2005)

# Create training and test data frames.
# Use the logical vector to subset rows
train <- Smarket[train_logic, ]
test  <- Smarket[!train_logic, ] # '!' means NOT (observations 2005 and later)
```

---

# 4. Linear Regression

## Fit a linear regression model

```{r linear-regression}
# Fit a linear regression model.
fit <- lm(Salary ~ ., data = train)

# Display the model summary.
summary(fit)
```

## Predict using linear regression

```{r linear-predict}
# Use the fitted model to predict on the test data.

pred <- predict(fit, newdata = test)
summary(pred)

# Calculate Test Mean Squared Error (MSE)
# This is the average of the squared differences between predicted and actual.
test_mse <- mean((pred - test$Salary)^2)

```

## Confidence intervals for coefficients

```{r linear-confint}
# Find confidence intervals for the regression coefficients.
# 'level' specifies the confidence level (default is 0.95).
confint(fit, level = 0.95)
```

## Diagnostic plots

```{r linear-diagnostics}
# Create diagnostic plots for the fitted linear regression model.

par(mfrow = c(2, 2))
plot(fit)

```

---

# 5. Bootstrap

## Bootstrap sample

Use this when you want to sample with replacement from the original dataset.

```{r bootstrap-sample}
# Sample row indices with replacement.
# 'replace = TRUE' is what makes it a bootstrap sample.
boot_indices <- sample(1:nrow(Hitters), nrow(Hitters), replace = TRUE)

# Create a bootstrap sample using those row indices.
boot_data <- Hitters[boot_indices, ]

```

## Bootstrap standard error of a statistic

Example: bootstrap standard error of the mean.

```{r bootstrap-se}
# Set a seed for reproducibility.
set.seed(1)

# Find the sample size.
n <- nrow(Hitters)

# Choose the number of bootstrap samples (B).
B <- 1000

# Create an empty vector to store the bootstrap statistics.
boot_stats <- numeric(B)

# Use a loop to create bootstrap samples and compute the statistic each time.
# Example statistic: Mean of Salary
for (i in 1:B) {
  indices <- sample(1:n, n, replace = TRUE)
  boot_sample <- Hitters[indices, ]
  boot_stats[i] <- mean(boot_sample$Salary, na.rm = TRUE)
}

# Compute the bootstrap standard error.
# The standard error is simply the standard deviation of the bootstrap distribution.
boot_se <- sd(boot_stats)
```

## Analytical standard error of the sample mean

```{r analytic-se}
# Compute the analytical standard error of the sample mean.
s <- sd(Hitters$Salary, na.rm = TRUE)
n <- sum(!is.na(Hitters$Salary)) # number of non-missing observations
analytic_se <- s / sqrt(n)
```

---

# 6. Logistic Regression

Use this when the response is binary, such as `Yes/No`.

## Fit logistic regression

```{r logistic-fit}

n = nrow(Smarket)
index <- sample(1:n,n/2) 

train_data <- Smarket[index, ]
test_data <- Smarket[-index, ]
# Fit a logistic regression model using glm().
# 'family = "binomial"' is essential to tell R to perform logistic regression.
glm.fit <- glm(Direction ~ Lag1 + Lag2, data = train_data, family = "binomial")

# Display the model summary.
# Look for coefficients with p-values < 0.05 to identify significant predictors.
summary(glm.fit)
```

## Predict probabilities

```{r logistic-prob}
# Predict probabilities on the test data.
# 'type = "response"' ensures R returns probabilities P(Y=1|X) instead of log-odds.
glm.probs <- predict(glm.fit, newdata = test_data, type = "response")
```

## Convert probabilities to class labels

```{r logistic-class}
# Start by assigning all observations to the baseline/negative class.
# Use contrasts(data$variable) to see which level is 0 and which is 1.
glm.pred <- rep("Down", nrow(test_data))

# Change predictions to the positive class when the probability is above the threshold.
# Standard threshold is 0.5.
glm.pred[glm.probs > 0.5] <- "Up"
```

## Confusion matrix and test error

```{r logistic-error}
# Create a confusion matrix.
# Format: table(Predicted, Actual)
conf_matrix <- table(glm.pred, test_data$Direction)
print(conf_matrix)

# Compute the test error rate.
# The mean of the logical vector where predictions do not match the actual values.
test_error <- mean(glm.pred != test_data$Direction)
cat("Test Error Rate:", test_error)

# Accuracy is simply (1 - test_error)
accuracy <- mean(glm.pred == test_data$Direction)
```

---

# 7. LDA and QDA

Use these for classification problems.

## LDA

Use LDA when predictors are approximately normally distributed within each class and covariance matrices are assumed equal across classes.

```{r lda}
# Fit an LDA model.
lda.fit <- lda(Direction ~ Lag1 + Lag2, data = train_data)

# Make predictions on the test data.
# Note: You do not need type="response" here.
lda.pred <- predict(lda.fit, newdata = test_data)

# Extract predicted class labels.
lda.class <- lda.pred$class

# Extract posterior probabilities.
# This returns a matrix where each column is a class (e.g., "Down", "Up").
lda.post <- lda.pred$posterior

# Create a confusion matrix.
table(lda.class, test_data$Direction)

# Compute accuracy.
mean(lda.class == test_data$Direction)

# Compute the misclassification rate.
mean(lda.class != test_data$Direction)
```

## Change classification threshold for LDA

Use this when you want to make the classifier more conservative. For example, predict `Up` only if the posterior probability of `Up` is larger than 0.52.

```{r lda-threshold}
# Start by assigning all observations to the baseline class (e.g., "Down").
lda.pred_adj <- rep("Down", nrow(test_data))

# Change predictions only when the posterior probability of the second class 
# (e.g., "Up") is larger than the chosen threshold.
# Check lda.pred$posterior to see which column index corresponds to "Up".
lda.pred_adj[lda.pred$posterior[, "Up"] > 0.52] <- "Up"

# Create a confusion matrix with the new threshold.
table(lda.pred_adj, test_data$Direction)

# Compute accuracy.
mean(lda.pred_adj == test_data$Direction)

# Compute the misclassification rate.
mean(lda.pred_adj != test_data$Direction)
```

## QDA

Use QDA when covariance matrices are allowed to differ across classes.

```{r qda}
# Fit a QDA model.
qda.fit <- qda(Direction ~ Lag1 + Lag2, data = train_data)

# Make predictions on the test data.
qda.pred <- predict(qda.fit, newdata = test_data)

# Extract predicted class labels.
qda.class <- qda.pred$class

# Extract posterior probabilities.
qda.post <- qda.pred$posterior

# Create a confusion matrix.
table(qda.class, test_data$Direction)

# Compute accuracy.
mean(qda.class == test_data$Direction)

# Compute the misclassification rate.
mean(qda.class != test_data$Direction)
```

---

# 8. Naive Bayes

Use Naive Bayes for classification when you assume predictors are conditionally independent within each class.

## Naive Bayes with normal density assumption

```{r naive-bayes}
# Fit a Naive Bayes model.
# Assumes quantitative predictors follow a Normal distribution within each class.
nb.fit <- naiveBayes(Direction ~ Lag1 + Lag2, data = train_data)

# Predict class labels on the test data.
nb.class <- predict(nb.fit, newdata = test_data)

# Predict class probabilities on the test data.
# type = "raw" returns the posterior probabilities for each class.
nb.probs <- predict(nb.fit, newdata = test_data, type = "raw")

# Create a confusion matrix.
table(nb.class, test_data$Direction)

# Compute accuracy.
mean(nb.class == test_data$Direction)
```

## Naive Bayes with kernel density estimation

Use this when you do not want to force the normal density assumption.

```{r naive-bayes-kernel}
# Fit a Naive Bayes model using kernel density estimation.
# usekernel = TRUE allows for a non-parametric estimate of the density.
nb.kernel.fit <- NaiveBayes(Direction ~ Lag1 + Lag2, data = train_data, usekernel = TRUE)

# Predict on the test data.
nb.kernel.pred <- predict(nb.kernel.fit, newdata = test_data)

# Extract predicted class labels.
nb.kernel.class <- nb.kernel.pred$class

# Extract posterior probabilities.
nb.kernel.post <- nb.kernel.pred$posterior

# Create a confusion matrix.
table(nb.kernel.class, test_data$Direction)

# Compute accuracy.
mean(nb.kernel.class == test_data$Direction)
```

---

# 9. KNN

Use KNN for classification when prediction is based on nearby observations.

## Standardize predictors

Important because KNN is distance-based.

```{r knn-standardize}
# Standardize the predictor variables so they have mean 0 and stdev 1.
# This prevents predictors with large scales from dominating the distance.
standardized.X <- scale(Smarket[, c("Lag1", "Lag2")])

# Verify variance is 1 for all columns
var(standardized.X[, 1])


# 1. Create the logical vector (TRUE for training years, FALSE for test years)
train_logic <- (Smarket$Year < 2005)

# 2. Now the KNN setup will work
train.X <- standardized.X[train_logic, ]
test.X <- standardized.X[!train_logic, ]

train.Direction <- Smarket$Direction[train_logic]
test.Direction <- Smarket$Direction[!train_logic]
```

## Train/test setup for KNN

```{r knn-setup}
# Create the training predictor matrix.
train.X <- standardized.X[train_logic, ]

# Create the test predictor matrix.
test.X <- standardized.X[!train_logic, ]

# Create the training response vector.
train.Direction <- Smarket$Direction[train_logic]

# Create the test response vector.
test.Direction <- Smarket$Direction[!train_logic]
```

## Fit KNN

```{r knn-fit}
# Set a seed for reproducibility (knn handles ties randomly).
set.seed(1)

# Fit KNN and predict class labels for the test data in one step.
# k is the number of nearest neighbors.
knn.pred <- knn(train.X, test.X, train.Direction, k = 1)
```

## Evaluate KNN

```{r knn-evaluate}
# Create a confusion matrix.
table(knn.pred, test.Direction)

# Compute accuracy.
mean(knn.pred == test.Direction)

# Compute the misclassification rate.
mean(knn.pred != test.Direction)
```

## Choose K using 5-fold cross-validation

```{r knn-cv}
# Set a seed.
set.seed(1)

# Create 5 folds from the training response.
folds <- createFolds(train.Direction, k = 5)

# Create a list of K values to try.
ks <- c(1, 3, 5, 7, 9, 11, 15, 20)

# Create an empty vector to store cross-validation accuracy for each K.
cv_accuracies <- numeric(length(ks))

# Loop over each K value.
for (i in 1:length(ks)) {
  fold_accuracies <- numeric(5)
  
  # For each K, loop over the 5 folds.
  for (j in 1:5) {
    # For each fold, create a validation set and a training set.
    val_idx <- folds[[j]]
    
    # Fit KNN on the training folds and predict on the validation fold.
    pred <- knn(train = train.X[-val_idx, ], 
                test = train.X[val_idx, ], 
                cl = train.Direction[-val_idx], 
                k = ks[i])
    
    fold_accuracies[j] <- mean(pred == train.Direction[val_idx])
  }
  
  # Store the average validation accuracy for each K.
  cv_accuracies[i] <- mean(fold_accuracies)
}

# Choose the K value with the highest cross-validation accuracy.
best_k <- ks[which.max(cv_accuracies)]
cat("Best K:", best_k)
```

---

# 10. ROC Curve and AUC

Use this when you want to evaluate a classifier across different thresholds.

```{r roc-curve}
# Create a prediction object using predicted probabilities and true class labels.
# Note: 'glm.probs' should be the numeric probabilities from predict(..., type="response")
# 'test_data$Direction' is the actual true class.
pred_obj <- prediction(glm.probs, test_data$Direction)

# Create the ROC performance object.
# "tpr" is True Positive Rate (Sensitivity), "fpr" is False Positive Rate (1 - Specificity)
roc_perf <- performance(pred_obj, measure = "tpr", x.measure = "fpr")

# Plot the ROC curve.
plot(roc_perf, colorize = TRUE, main = "ROC Curve")
abline(a = 0, b = 1, lty = 2, col = "gray") # Add diagonal line for random guessing
```

## ROC curve with cutoff labels

```{r roc-cutoffs}
# Plot the ROC curve with cutoff labels.
plot(roc_perf, colorize = TRUE, print.cutoffs.at = seq(0, 1, by = 0.1), 
     text.adj = c(-0.2, 1.7), main = "ROC Curve with Cutoffs")
```

## Compute AUC

```{r auc}
# Compute the AUC.
auc_perf <- performance(pred_obj, measure = "auc")

# Extract the AUC value from the nested list structure.
auc_value <- auc_perf@y.values[[1]]
cat("AUC:", auc_value)
```

---

# 11. Classification Trees

Use this when the response is categorical.

## Fit a classification tree

```{r classification-tree}

n = nrow(Carseats)
index <- sample(1:n,n/2) 

train_data <- Carseats[index, ]
test_data <- Carseats[-index, ]

High <- factor(ifelse(Carseats$Sales <= 8, "No", "Yes"))

# Fit a classification tree.
# Ensure the response variable is a factor for a classification tree.
tree.fit <- tree(High ~ . - Sales, data = train_data)

# Display a summary of the tree.
# Shows variables used, number of terminal nodes, and training error rate.
summary(tree.fit)

# Plot the tree.
plot(tree.fit)

# Add text labels to the tree.
# 'pretty = 0' includes category names for qualitative predictors.
text(tree.fit, pretty = 0)
```

## Predict class labels

```{r classification-tree-predict}
# Predict class labels on the test data.
# 'type = "class"' is required to get the category labels instead of probabilities.
tree.pred <- predict(tree.fit, newdata = test_data, type = "class")
```

## Test classification error

```{r classification-tree-error}
# Create a confusion matrix.
table(tree.pred, test_data$High)

# Compute the test error rate.
mean(tree.pred != test_data$High)
```

## Prune a classification tree

```{r classification-tree-prune}
# Set a seed for reproducibility of the cross-validation folds.
set.seed(3)

# Use cross-validation to choose the tree size.
# 'FUN = prune.misclass' tells R to use the misclassification rate to guide pruning.
cv.tree_obj <- cv.tree(tree.fit, FUN = prune.misclass)

# Plot cross-validation error (dev) against tree size (size).
plot(cv.tree_obj$size, cv.tree_obj$dev, type = "b", xlab = "Tree Size", ylab = "CV Error")

# Find the best tree size (the one with the lowest CV error).
best_size <- cv.tree_obj$size[which.min(cv.tree_obj$dev)]

# Prune the tree using the best size.
pruned.tree <- prune.misclass(tree.fit, best = best_size)

# Plot the pruned tree.
plot(pruned.tree)

# Add text labels to the pruned tree.
text(pruned.tree, pretty = 0)
```

---

# 12. Regression Trees

Use this when the response is quantitative.

## Fit a regression tree

```{r regression-tree}

n = nrow(Hitters)
index <- sample(1:n,n/2) 

train_data <- Hitters[index, ]
test_data <- Hitters[-index, ]

# Fit a regression tree.
# Note: Ensure the response (Salary) is numeric.
tree.fit <- tree(Salary ~ ., data = train_data)

# Display a summary of the tree.
# For regression, this shows variables used and the Residual Mean Deviance (MSE).
summary(tree.fit)

# Plot the tree.
plot(tree.fit)

# Add text labels to the tree.
text(tree.fit, pretty = 0)
```

## Predict and calculate test MSE

```{r regression-tree-mse}
# Predict on the test data.
# Note: type = "class" is NOT used here; we want numeric values.
tree.pred <- predict(tree.fit, newdata = test_data)

# Compute the test MSE.
test_mse <- mean((tree.pred - test_data$Salary)^2)
cat("Regression Tree Test MSE:", test_mse)
```

## Number of terminal nodes

```{r terminal-nodes}
# Find the number of terminal nodes in the tree.
summary(tree.fit)$size
```

## Prune a regression tree

```{r regression-tree-prune}
# Set a seed.
set.seed(1)

# Use cross-validation to choose the tree size.
# For regression, we do NOT use FUN = prune.misclass. 
# The default is to use deviance (RSS).
cv.tree_obj <- cv.tree(tree.fit)

# Plot cross-validation error against tree size.
plot(cv.tree_obj$size, cv.tree_obj$dev, type = "b", xlab = "Tree Size", ylab = "CV RSS")

# Find the best tree size.
best_size <- cv.tree_obj$size[which.min(cv.tree_obj$dev)]

# Prune the tree using the best size.
pruned.tree <- prune.tree(tree.fit, best = best_size)

# Predict using the pruned tree.
pruned.pred <- predict(pruned.tree, newdata = test_data)

# Compute the test MSE for the pruned tree.
pruned_mse <- mean((pruned.pred - test_data$Salary)^2)
cat("Pruned Tree Test MSE:", pruned_mse)
```

---

# 13. Bagging

Bagging is a special case of random forests where `mtry = p`, meaning all predictors are considered at each split.

## Fit bagging model

```{r bagging}
# Find the number of predictors (p). 
# Subtract 1 because 'Salary' is the response, not a predictor.
p <- ncol(train_data) - 1

# Fit a bagging model by setting mtry equal to the number of predictors.
# 'importance = TRUE' allows us to view variable importance later.
set.seed(1)
bag.fit <- randomForest(Salary ~ ., data = train_data, mtry = p, importance = TRUE)
```

## Predict and calculate MSE

```{r bagging-mse}
# Predict on the test data.
bag.pred <- predict(bag.fit, newdata = test_data)

# Compute the test MSE.
bag_mse <- mean((bag.pred - test_data$Salary)^2)
cat("Bagging Test MSE:", bag_mse)
```

## Variable importance

```{r bagging-importance}
# View variable importance in a table.
# %IncMSE: How much the MSE increases if we remove that variable.
# IncNodePurity: Total decrease in node impurity (RSS) from splitting on that variable.
importance(bag.fit)

# Plot variable importance for a quick visual.
varImpPlot(bag.fit)
```

---

# 14. Random Forests

Random forests improve bagging by using only a random subset of predictors at each split.

## Fit random forest model

```{r random-forest}
# Fit a random forest model. 
# Typical default for regression is p/3. Let's try mtry = 6.
set.seed(1)
rf.fit <- randomForest(Salary ~ ., data = train_data, mtry = 6, importance = TRUE)
```

## Predict and calculate MSE

```{r random-forest-mse}
# Predict on the test data.
rf.pred <- predict(rf.fit, newdata = test_data)

# Compute the test MSE.
rf_mse <- mean((rf.pred - test_data$Salary)^2)
cat("Random Forest Test MSE:", rf_mse)
```

## Choose `mtry` in a data-driven way

```{r random-forest-mtry}
# Find the number of predictors.
p <- ncol(train_data) - 1

# Create possible mtry values (from 1 to p).
mtry_values <- 1:p

# Create an empty vector to store test errors.
test_errors <- numeric(p)

# Loop over possible mtry values.
for (m in mtry_values) {
  set.seed(1)
  # For each mtry value, fit a random forest model.
  rf.tmp <- randomForest(Salary ~ ., data = train_data, mtry = m)
  
  # Predict on the test data.
  pred.tmp <- predict(rf.tmp, newdata = test_data)
  
  # Store the test MSE.
  test_errors[m] <- mean((pred.tmp - test_data$Salary)^2)
}

# Choose the mtry value with the lowest test MSE.
best_mtry <- which.min(test_errors)
cat("Optimal mtry:", best_mtry)

# Plot the results to see the trade-off
plot(mtry_values, test_errors, type = "b", xlab = "mtry", ylab = "Test MSE")
```

## Variable importance

```{r random-forest-importance}
# View variable importance.
importance(rf.fit)

# Plot variable importance.
varImpPlot(rf.fit)
```

---

# 15. Boosting for Regression Trees

Boosting fits trees sequentially. Each new tree tries to improve the current model.

## Fit boosted regression trees

```{r boosting}
# Fit a boosted regression tree model.
# 'distribution = "gaussian"' is for regression; use "bernoulli" for binary classification.
# 'n.trees = 5000' is the number of trees.
# 'interaction.depth = 4' limits the number of splits in each tree.
set.seed(1)
boost.fit <- gbm(Salary ~ ., data = train_data, distribution = "gaussian", 
                 n.trees = 5000, interaction.depth = 4)
```

## Variable importance

```{r boosting-importance}
# View variable importance for the boosting model.
# This produces a summary table and a bar plot by default.
summary(boost.fit)
```

## Partial dependence plots

```{r partial-dependence}
# Create partial dependence plots for selected predictors.
# Example: 'LStat' and 'RM' (if using Boston data) or 'CAtBat' and 'CHits' (if using Hitters).
plot(boost.fit, i = "CAtBat")
plot(boost.fit, i = "CHits")
```

## Predict and calculate test MSE

```{r boosting-mse}
# Predict on the test data using the chosen number of trees.
# You MUST specify n.trees in the predict function for gbm.
boost.pred <- predict(boost.fit, newdata = test_data, n.trees = 5000)

# Compute the test MSE.
boost_mse <- mean((boost.pred - test_data$Salary)^2)
cat("Boosting Test MSE:", boost_mse)
```

## Try a different shrinkage value

```{r boosting-shrinkage}
# Fit another boosted regression tree model with a higher shrinkage (learning rate).
boost.fit2 <- gbm(Salary ~ ., data = train_data, distribution = "gaussian", 
                  n.trees = 5000, interaction.depth = 4, shrinkage = 0.2, verbose = F)

# Predict on the test data.
boost.pred2 <- predict(boost.fit2, newdata = test_data, n.trees = 5000)

# Compute the test MSE.
boost_mse2 <- mean((boost.pred2 - test_data$Salary)^2)
cat("Boosting (Shrinkage=0.2) Test MSE:", boost_mse2)
```

---

# 16. Error and Accuracy

## Regression mean squared error

Use this when the response is quantitative.

```{r mse}
# Note: Assumes you have already defined 'test_data' and 'pred' (numeric predictions)
# Formula: mean((actual - predicted)^2)
mse <- mean((test_data$Salary - pred)^2)
cat("Test MSE:", mse)
```

## Classification accuracy

Use this when the response is categorical.

```{r accuracy}
# 1. Setup Data and Model
library(ISLR2)
train_logic <- (Smarket$Year < 2005)
train_data  <- Smarket[train_logic, ]
test_data   <- Smarket[!train_logic, ]

# 2. Fit and Predict
glm.fit   <- glm(Direction ~ Lag1 + Lag2, data = train_data, family = "binomial")
glm.probs <- predict(glm.fit, newdata = test_data, type = "response")

# 3. Create pred.class (Converting probabilities to labels)
pred.class <- rep("Down", nrow(test_data))
pred.class[glm.probs > 0.5] <- "Up"
pred.class <- as.factor(pred.class)

# 4. Compute accuracy
# Compare predicted labels to actual labels in test_data
accuracy <- mean(test_data$Direction == pred.class)
cat("Accuracy:", accuracy)
```

## Classification error rate

```{r error-rate}
# Method 1: Mean of mismatches
error_rate <- mean(test_data$Direction != pred.class)

# Method 2: Complement of accuracy
error_rate <- 1 - accuracy
cat("Error Rate:", error_rate)
```

## Confusion matrix

```{r confusion-matrix}
# Create a confusion matrix.
# table(Predicted, Actual)
conf_matrix <- table(pred.class, test_data$Direction)
print(conf_matrix)
```

## Precision

Use this when the positive class is rare and you care about how many predicted positives are truly positive.

```{r precision}
# Create a confusion matrix object
cm <- table(pred.class, test_data$Direction)

# Compute precision for the positive class (e.g., "Up")
# cm["Predicted", "Actual"]
precision <- cm["Up", "Up"] / sum(cm["Up", ])
cat("Precision:", precision)
```

---

# 17. Common Model Comparison Table

| Method | Response Type | Main Function | Prediction Function | Main Evaluation |
|---|---|---|---|---|
| Linear regression | Quantitative | `lm()` | `predict()` | MSE |
| Logistic regression | Binary | `glm(..., family="binomial")` | `predict(..., type="response")` | Accuracy, error, ROC/AUC |
| LDA | Categorical | `lda()` | `predict()` | Accuracy, error |
| QDA | Categorical | `qda()` | `predict()` | Accuracy, error |
| Naive Bayes | Categorical | `naiveBayes()` | `predict()` | Accuracy, error |
| KNN | Categorical | `knn()` | `knn()` directly predicts | Accuracy, error, precision |
| Classification tree | Categorical | `tree()` | `predict(..., type="class")` | Accuracy, error |
| Regression tree | Quantitative | `tree()` | `predict()` | MSE |
| Bagging | Both | `randomForest(..., mtry=p)` | `predict()` | MSE or accuracy |
| Random forest | Both | `randomForest()` | `predict()` | MSE or accuracy |
| Boosting | Usually quantitative here | `gbm()` | `predict(..., n.trees=...)` | MSE |

---

# 18. Quick Decision Guide

| If the problem asks you to... | Use this |
|---|---|
| Estimate uncertainty by resampling | Bootstrap |
| Predict a quantitative response with a straight-line model | Linear regression |
| Predict a binary response using probabilities | Logistic regression |
| Compare thresholds | ROC curve / AUC |
| Classify using normal class distributions | LDA / QDA |
| Classify using Bayes theorem and conditional independence | Naive Bayes |
| Classify based on nearby observations | KNN |
| Build an interpretable tree | Classification or regression tree |
| Reduce variance of trees by averaging many trees | Bagging |
| Decorrelate trees by randomly selecting predictors | Random forest |
| Fit trees sequentially to improve prediction | Boosting |

---
