---
title: "HW9"
author: "Nathan Krieger"
date: "2026-04-08"
output: pdf_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```


# Problem 1: Bootstrap

We will work with the Boston housing data set from the ISLR2 library. The first part of
this problem is similar to our in-class activity on Bootstrap.

## (a)

Based on this data set, provide an estimate for the population mean of medv. Call this
estimate ˆu. 


```{r}

library(ISLR2)

n <- nrow(Boston)

# (1) Estimate the population mean of medv
mu_hat <- mean(Boston$medv)
mu_hat


```



## (b) 

Provide an estimate of the standard error of ˆu using an analytical formula. Interpret
this result.

```{r}

se_analytic <- sd(Boston$medv) / sqrt(n)
se_analytic

```

## (c) 

Now estimate the standard error of ˆu using the bootstrap. How does this compare to
your answer from part (b)?

```{r}

B <- 2000

boot_mean <- numeric(B) # (something to store boot mean, how many do we have ?)

for (b in 1:B) {
  boot_index <- sample(1:n, size = n, replace = TRUE)
  boot_sample <- Boston$medv[boot_index]
  
  boot_mean[b] <- mean(boot_sample)
}

se_boot <- sd(boot_mean)
se_boot

c(bootstrap_SE = se_boot, analytic_SE = se_analytic)

```

## (d) 

Using bootstrap, provide a 95% confidence interval for the mean of medv. Compare it
to the results using analytical formulas.


``` {r}

# 95% bootstrap percentile confidence interval
boot_ci <- quantile(boot_mean, c(0.025, 0.975))

t_score <- qt(0.975, df = n - 1)
analytic_ci <- c(mu_hat - t_score * se_analytic, mu_hat + t_score * se_analytic)

# Results
list(Bootstrap_CI = boot_ci, Analytic_CI = analytic_ci)

```

## (e) 

Based on this data set, provide an estimate ˆumed for the median value of medv.


``` {r}

med_hat <- median(Boston$medv)
med_hat

```

## (f) 

We would like to estimate the standard error of ˆumed. Since there is no simple formula
for computing the standard error of the median, use bootstrap. Comment on your
findings.

``` {r}

set.seed(1)
B <- 2000
boot_median <- numeric(B)

for (b in 1:B) {
  boot_index <- sample(1:n, size = n, replace = TRUE)
  boot_sample <- Boston$medv[boot_index]
  boot_median[b] <- median(boot_sample)
}

se_median_boot <- sd(boot_median)
se_median_boot

```

# Problem 2: Bootstrap Investments (practice)

Suppose we wish to invest a fixed sum of money into two financial assets that yield
returns of X and Y , respectively. We will invest a fraction a of our money in X and invest
the remaining 1 - a in Y . We want to find the value of a that minimizes the total risk of
our investment. One can show that the value that minimizes the risk is given by

[FORMULA]

which needs to be estimated from the data.

The Portfolio data set in the ISLR2 package contains data for 100 pairs of stock returns.
The R script alpha fn.R will compute ˆa for you for this data set. Use bootstrap to quan-
tify the accuracy of our estimate of ˆa. In other words, estimate the standard error using
bootstrap.



# Problem 3: Email Spam

We will use a well-known dataset to practice classification. You can find it here: https://
archive.ics.uci.edu/ml/datasets/Spambase. Read the attribute information and down-
load the dataset onto your computer. To load this data into R, use the following code:

```{r}
spam = read.csv("spambase.data", header = FALSE)
```

The last column of the spam data set, called V58, denotes whether the e-mail was considered
spam (1) or not (0).

## (a) 

What proportion of emails are classified as spam and what proportion of emails are
non-spam?

## (b) 

Carefully split the data into training and testing sets. Check to see that the proportions
of spam vs. non-spam in your training and testing sets are similar to what you observed
in part (a). Report those proportions here.

## (c) 

Fit a logistic regression model on the training set and evaluate it on the test set. Use
all variables as predictors (except for V58) to train the model. Use the predict()
function to predict the probability that an email in our data set will be spam or not.
Print the first ten predicted probabilities from your test set.

## (d) 

We can convert these probabilities into labels. If the predicted probability is greater
than 0.5, then we predict the email is spam ( ˆYi = 1); otherwise it is not spam ( ˆYi = 0).
Create a confusion matrix based on your results. What is the overall misclassification
rate? Break this down and report the false negative rate and false positive rate.

## (e) 

What type of mistake do we think is more critical here: reporting a meaningful email as
spam or a spam email as meaningful? How can we adjust our classifier to accommodate
this?

